---
title: "Data607 - Week 5 Assignment"
author: "Peter Kowalchuk"
date: "9/23/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Tidying and Transforming Data

```{r}
library(tidyr)
library(dplyr)
```


###1) CSV Data

Data was entered in a csv file called flightsInfo.csv using the wide format given

###2) Read CSV into R, tidy and transform data

CSV read into R
```{r}
dfWide<-read.csv(file="flightsInfo.csv")
dfWide
tbWide<-tbl_df(dfWide)
tbWide
```

We reshape the data by going from a wide table to a tidy table with variables as columns and observations as rows

```{r}
tbTidy<-gather(dfWide,"Destinations","Flights",3:7)
tbTidy
```

We then group the data by airline and delayed fligths

```{r}
tbTidyGroupAir<-group_by(tbTidy,Airline,Delayed)
tbTidyGroupAir
```

###3) Analysis comparing arrival delays for both airlines

Using the tidy table with groups we can compare both airlines arrival times performing a summary. From here we can see how AM West has more delayed flights than Alaska, but also has more on time. Overall AM West has more flights than Alaska, so a better way to do this comparison is using ratios.

```{r}
summarise(tbTidyGroupAir,totalFligths=sum(Flights))
```

Before continuing, we can point out how all these operations can be done using pipies, which will be the method used going forward

```{r}
dfWide %>% tbl_df() %>% gather("Destinations","Flights",3:7) %>% group_by(Airline) %>% summarise(totalFligths=sum(Flights))
dfWide %>% tbl_df() %>% gather("Destinations","Flights",3:7) %>% group_by(Airline,Delayed) %>% summarise(totalFligths=sum(Flights))
```

To calculate ratios we can mutate our final table from before to add a ratio column

```{r}
dfWide %>% tbl_df() %>% gather("Destinations","Flights",3:7) %>% group_by(Airline,Delayed) %>% summarise(totalFligths=sum(Flights)) %>% mutate(Ratio=totalFligths/sum(totalFligths))
```

From the ratio shown above we discover that Alaska has slightly more delayed flights, with 13.3% of its flights being delayed against 10.9% of AM West flights being delayed.

###1) Data from SQL databse

Instead of loading the data into R from a csv file, we could have also created a table in MySQL and imported it into R. 

```{r}
library("RMySQL")
mydb<-dbConnect(MySQL(),user='root',password='123',dbname='flightDelays',host='localhost')
dbListTables(mydb)
dbListFields(mydb,'tableWide')
tableWideDataQuery<-dbSendQuery(mydb,'select * from tableWide')
dfTableDbWide<-fetch(tableWideDataQuery,n=-1)
dfTableDbWide
tbWide<-tbl_df(dfTableDbWide)
tbWide
```

The resulting tbWide table is the same table we obtained from the CSV import. The same analysis can be performed using this table as the starting point.
